|

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): use clearpage for page break
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): specify per-format affiliations
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: |
  Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-mathphys-num
    classoption: 
      - lineno
    cite-method: natbib
    link-citations: true
    toc: true
    toc-title: Contents
    number-sections: true
    keep-tex: true
    equal-margins: false
    include-in-header: header.tex
  html:
    code-fold: true
    html-math-method: mathjax
  docx:
    reference-doc: styles.docx
    cite-method: citeproc
    filters:
      - authors-block
csl: _extensions/sciexp/nature/csl/springer-mathphys-brackets.csl
affiliations:
  - name: Massachusetts General Hospital Research Institute, Charlestown, MA
    id: 1
    department: Molecular Pathology Unit
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Harvard Medical School, Charlestown, MA
    id: 2
    department: Massachusetts General Hospital Cancer Center
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Massachusetts General Hospital, Boston, MA
    id: 3
    department: Center for Regenerative Medicine
    city: Boston
    state: MA
    postal-code: 02114
  - name: Harvard University, Cambridge, MA
    id: 4
    department: Harvard Stem Cell Institute
    city: Cambridge
    state: MA
    postal-code: 02139
  - name: Cambridge, MA
    id: 5
    department: Broad Institute of MIT and Harvard
    city: Cambridge
    state: MA
    postal-code: 02142
  - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
    id: 6
    department: |
      Brain and Mind Institute, School of Life Sciences, 
      École Polytechnique Fédérale de Lausanne (EPFL)
    city: Lausanne
    state: Switzerland
    postal-code: 1015
  - name: Cambridge, MA
    id: 7
    department: Basis Research Institute
    city: Cambridge
    state: MA
    postal-code: 02142
authors:
  - name: Qian Qin
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
      - ref: 5
    # attributes:
    #   equal-contributor: true
  - name: Eli Bingham
    affiliations:
      - ref: 5
      - ref: 7
  - name: Gioele La Manno
    affiliations:
      - ref: 6
    email: gioele.lamanno@epfl.ch
    attributes:
      corresponding: true
  - name: David M. Langenau
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
    email: dlangenau@mgh.harvard.edu
    attributes:
      corresponding: true
  - name: Luca Pinello
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 5
    email: lpinello@mgh.harvard.edu
    attributes:
      corresponding: true
abstract: |
  Single-cell RNA Velocity has dramatically advanced our ability to model
  cellular differentiation and cell fate decisions. However, current
  preprocessing choices and model assumptions often lead to errors in assigning
  developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian,
  generative, and multivariate RNA Velocity model to estimate the uncertainty of
  cell future states. This approach models raw sequencing counts with the
  synchronized cell time across all expressed genes to provide quantifiable and
  improved information on cell fate choices and developmental trajectory
  dynamics.
keywords: [
  RNA velocity, 
  single-cell RNA sequencing, 
  probabilistic generative modeling, 
  cell fate, 
  developmental trajectory inference
]
bibliography: bibliography.bib
crossref:
  fig-prefix: Fig.
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

@@

Pyro-Velocity recasts the velocity estimation problem into a latent variable
posterior probability inference (\nameref{sec-methods}). The proposed model is
generative and fully Bayesian, with the different parameters considered as
latent random variables. Each parameter is assigned a prior distribution, and
the model is then conditioned on the raw data (both spliced and unspliced
molecule raw counts) to estimate posterior probability distributions and to
enable uncertainty estimation of velocity. Central to the Pyro-Velocity model is
a *shared latent time* $(t)$: a global latent variable placing each cell at a
particular instant of the modeled dynamical process. From this latent time, the
expectation of spliced and unspliced RNA levels for each gene can be computed
knowing gene-specific kinetics parameters (transcriptional $(\alpha)$, splicing
$(\beta)$, and degradation rates ($\gamma$) and two gene-specific switching
times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
integrating the RNA velocity differential equations analogously to what proposed
in the dynamical RNA velocity model by Bergen et al.
[@Bergen2020-pj;@Li2021-qa]. However, we innovate the structure and formulation
of the latent time given that our model is set to jointly fit the entire raw
data matrix, not individual genes. We approach the task of estimating the
posterior distribution by restricting the search within a family of a
variational distribution approximating the true posterior, and taking advantage
of the automatic differentiation variational inference (ADVI) capabilities of
Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
\nameref{sec-methods}). Notably, our generative Bayesian model enables to sample
from the posterior distributions of the latent and dependent variables, allowing
uncertainty estimation of phase portraits, high dimensional velocity vector
field, and shared latent time. Around this model, we built several
visualizations to assess uncertainty and functions to recover putative cell fate
priming genes and to investigate how the denoised gene expression recapitulates
known cell-type annotations and their developmental order ([@fig-model]b,
\nameref{sec-methods}). Although several RNA velocity methods
[@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
recently proposed, and many of them have not been peer-reviewed, here we
showcase the advantages and innovations of Pyro-Velocity by comparing it with
the current state-of-the-art dynamical RNA velocity model proposed by Bergen et
al. [@Bergen2020-pj], and implemented in scVelo.

@@

>>>>

---
title: |
  Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data
toc: true
toc-title: Contents
number-sections: true
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-mathphys-num
    classoption: 
      - lineno
    cite-method: natbib
    link-citations: true
    keep-tex: true
    equal-margins: false
    include-in-header: header.tex
    affiliations:
      - name: Massachusetts General Hospital Research Institute, Charlestown, MA
        id: 1
        department: Molecular Pathology Unit
        city: Charlestown
        state: MA
        postal-code: 02129
      - name: Harvard Medical School, Charlestown, MA
        id: 2
        department: Massachusetts General Hospital Cancer Center
        city: Charlestown
        state: MA
        postal-code: 02129
      - name: Massachusetts General Hospital, Boston, MA
        id: 3
        department: Center for Regenerative Medicine
        city: Boston
        state: MA
        postal-code: 02114
      - name: Harvard University, Cambridge, MA
        id: 4
        department: Harvard Stem Cell Institute
        city: Cambridge
        state: MA
        postal-code: 02139
      - name: Cambridge, MA
        id: 5
        department: Broad Institute of MIT and Harvard
        city: Cambridge
        state: MA
        postal-code: 02142
      - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
        id: 6
        department: |
          Brain and Mind Institute, School of Life Sciences, 
          École Polytechnique Fédérale de Lausanne (EPFL)
        city: Lausanne
        state: Switzerland
        postal-code: 1015
      - name: Cambridge, MA
        id: 7
        department: Basis Research Institute
        city: Cambridge
        state: MA
        postal-code: 02142
  html:
    code-fold: true
    html-math-method: mathjax
    affiliations:
      - name: Molecular Pathology Unit, Massachusetts General Hospital Research Institute, Charlestown, MA
        id: 1
      - name: Massachusetts General Hospital Cancer Center, Harvard Medical School, Charlestown, MA
        id: 2
      - name: Center for Regenerative Medicine, Massachusetts General Hospital, Boston, MA
        id: 3
      - name: Harvard Stem Cell Institute, Harvard University, Cambridge, MA
        id: 4
      - name: Broad Institute of MIT and Harvard, Cambridge, MA
        id: 5
      - name: Laboratory of Neurodevelopmental Systems Biology, Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland
        id: 6
      - name: Basis Research Institute, Cambridge, MA
        id: 7
  docx:
    reference-doc: styles.docx
    cite-method: citeproc
    filters:
      - authors-block
    affiliations:
      - name: Molecular Pathology Unit, Massachusetts General Hospital Research Institute, Charlestown, MA
        id: 1
      - name: Massachusetts General Hospital Cancer Center, Harvard Medical School, Charlestown, MA
        id: 2
      - name: Center for Regenerative Medicine, Massachusetts General Hospital, Boston, MA
        id: 3
      - name: Harvard Stem Cell Institute, Harvard University, Cambridge, MA
        id: 4
      - name: Broad Institute of MIT and Harvard, Cambridge, MA
        id: 5
      - name: Laboratory of Neurodevelopmental Systems Biology, Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland
        id: 6
      - name: Basis Research Institute, Cambridge, MA
        id: 7
csl: _extensions/sciexp/nature/csl/springer-mathphys-brackets.csl
authors:
  - name: Qian Qin
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
      - ref: 5
    # attributes:
    #   equal-contributor: true
  - name: Eli Bingham
    affiliations:
      - ref: 5
      - ref: 7
  - name: Gioele La Manno
    affiliations:
      - ref: 6
    email: gioele.lamanno@epfl.ch
    attributes:
      corresponding: true
  - name: David M. Langenau
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
    email: dlangenau@mgh.harvard.edu
    attributes:
      corresponding: true
  - name: Luca Pinello
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 5
    email: lpinello@mgh.harvard.edu
    attributes:
      corresponding: true
abstract: |
  Single-cell RNA Velocity has dramatically advanced our ability to model
  cellular differentiation and cell fate decisions. However, current
  preprocessing choices and model assumptions often lead to errors in assigning
  developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian,
  generative, and multivariate RNA Velocity model to estimate the uncertainty of
  cell future states. This approach models raw sequencing counts with the
  synchronized cell time across all expressed genes to provide quantifiable and
  improved information on cell fate choices and developmental trajectory
  dynamics.
keywords: [
  RNA velocity, 
  single-cell RNA sequencing, 
  probabilistic generative modeling, 
  cell fate, 
  developmental trajectory inference
]
bibliography: bibliography.bib
crossref:
  fig-prefix: Fig.
  sec-prefix: Sec.
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

@@

Pyro-Velocity recasts the velocity estimation problem into a latent variable
posterior probability inference (@sec-methods). The proposed model is
generative and fully Bayesian, with the different parameters considered as
latent random variables. Each parameter is assigned a prior distribution, and
the model is then conditioned on the raw data (both spliced and unspliced
molecule raw counts) to estimate posterior probability distributions and to
enable uncertainty estimation of velocity. Central to the Pyro-Velocity model is
a *shared latent time* $(t)$: a global latent variable placing each cell at a
particular instant of the modeled dynamical process. From this latent time, the
expectation of spliced and unspliced RNA levels for each gene can be computed
knowing gene-specific kinetics parameters (transcriptional $(\alpha)$, splicing
$(\beta)$, and degradation rates ($\gamma$) and two gene-specific switching
times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
integrating the RNA velocity differential equations analogously to what proposed
in the dynamical RNA velocity model by Bergen et al.
[@Bergen2020-pj;@Li2021-qa]. However, we innovate the structure and formulation
of the latent time given that our model is set to jointly fit the entire raw
data matrix, not individual genes. We approach the task of estimating the
posterior distribution by restricting the search within a family of a
variational distribution approximating the true posterior, and taking advantage
of the automatic differentiation variational inference (ADVI) capabilities of
Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
@sec-methods). Notably, our generative Bayesian model enables to sample
from the posterior distributions of the latent and dependent variables, allowing
uncertainty estimation of phase portraits, high dimensional velocity vector
field, and shared latent time. Around this model, we built several
visualizations to assess uncertainty and functions to recover putative cell fate
priming genes and to investigate how the denoised gene expression recapitulates
known cell-type annotations and their developmental order ([@fig-model]b,
@sec-methods). Although several RNA velocity methods
[@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
recently proposed, and many of them have not been peer-reviewed, here we
showcase the advantages and innovations of Pyro-Velocity by comparing it with
the current state-of-the-art dynamical RNA velocity model proposed by Bergen et
al. [@Bergen2020-pj], and implemented in scVelo.

@@

\FloatBarrier
\clearpage
{{< pagebreak >}}
